### 2017-11-08 paper/response_to_reviewers.tex

\begin{enumerate}

\item{\textcolor{blue}{While I understand that the author is writing a package that has many diverse capabilities and is presented to be useful to a wide range of disciplines, it is too vague in too many places to be helpful in its current state.
For example, the line in the abstract “It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.”.
“Some properties” is too vague without any qualifiers to be useful. But I believe that with revision, taking into account the issues of clarity, this will be an important addition to the available analytical tools for Biologists.}}

I have changed the terms used in the abstract and proposed more specific examples for clarity:

\textit{It is then possible to measure divers aspects of the distribution of some observation (e.g. species) in this space. For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} line @@@

@@

The changes have also been made throughout the manuscript as generally suggested by all four reviewers.
%TODO: link to other clarity changes

\item{\textcolor{blue}{My main concern is that what disparity actually means is left to the reader’s imagination – it is not defined or discussed anywhere.
Disparity simply means lack of similarity or equality.
While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
%TODO improve disparity definition in the abstract
Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
Say what “specific aspect of it [multidimensional space]” you are referring to.
Also, the different definitions of disparity alluded to on lines 61-63 should be summarised.
Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}

%Precision up on what is disparity

As Reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
I feel that explaining this would be out of the scope of this software description paper.
However, I have tried to give specific examples each time I refer to disparity as a way to measure these ``specific aspect of the space'' and I have also added a clearer definition of what I mean by disparity throughout this manuscript.

I rewrote the part in the introducing disparity as follows:

\textit{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. “does group A occupies more space than group B?”). This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways. For example, one could use a metric based on the variance or the range of each axis of space (Wills, 2001; Ciampaglio et al., 2001) , a distance (e.g. Euclidean) measured between observations (Foote, 1993, 1996) , a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013) , or many more (e.g. Navarro, 2003).} lines @@@

@@

I have fixed these typos.

\item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}

% Define Procrustes variance

Procrustes variance = Sum of the diagonal covariance matrix between each elements divided by the number of elements.
This is appropriate when looking simply at the volume in the multidimensional space. Howver, does not work if we want to look at the distance between two groups or if we want to use another proxy than the procrustes variance for the space occupancy (e.g. the ellipsoid volume might and direction might be more appropriate in XXX case).

%From the manual: Procrustes variance is the same sum of the diagonal elements of the group covariance matrix divided by the number of observations in the group

%From Michael Collyer: As the author indicates, in geomorph we limit our disparity measure to Procrustes variance (same as the disparity metric used in dispRity) and hypothesis tests to use randomization tests that randomly permute residuals (not data) over many permutations to generate pairwise differences in disparity.

@@

\item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}

I have changed the sentence following the reviewer's suggestion.

\item{\textcolor{blue}{Line 116 – “Table ??” = should have been caught during proofreading. Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
\label{table_missing}

I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the currently implemented metrics description.
I would also like to underline that although not all usual disparity metrics are not represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.

For example, if one decides to use the relative ellipsoid hyper volume to the number of elements per group as a disparity metric (i.e. this metric could be used as a proxy for the average space occupancy per unit element - e.g. per taxon).
This would not be doable directly by using the implemented \texttt{ellipse.volume} function but would require the user to use a function like:\\
\texttt{dispRity(my\_data, metric = function(X) ellipse.volume(X)/nrows(X))}\\

\item{\textcolor{blue}{Lines 356-357 – This final sentence of the conclusion section is vague and does not add to the manuscript.}}
\label{remove_last_sentence}

I removed the last sentence.

@@

\item{\textcolor{blue}{Figure 1 – Procrustes is a proper noun and requires a capital p. Is the Procrustes matrix a “Procrustes distance matrix” or a “Procrustes residuals matrix”? i.e. is it the shape variables after a Procrustes superimposition (latter) or the distances between individuals after a Procrustes superimposition (former)? Both are suitable inputs for ordination.}}

I meant both matrices (as this reviewers rightly points out they are suitable for ordination).
I have made this more clear in the figure.
On a more general note, the package now allows users to input any matrix (although the package does not provide warranty that the provided matrix is meaningful to the user's specific analysis).

### 2017-11-08 paper/dispRity_paper_highlighted.tex
Updated responses

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
Once the multidimensional space is defined it is then possible to measure a specific aspect of it.
This specific measurement (hereafter referred to as a ``disparity metric'') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

@@

This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
\hl{The first argument is the matrix} defining the multidimensional space and a list of elements to group into different subsamples.
The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example, \hl{see below}) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).

>>>>

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep[][and references therein]{lloyd2016estimating}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.

\hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
This requires the definition of a proxy for space occupancy: the disparity metric}
\citep[or index;][]{Hopkins2017}
\hl{which can be measured in a multitude of ways.}
\hl{For example, one could use a metric based on the variance or the range of each axis of space}
\citep{Wills2001, Ciampaglio2001}
\hl{, a distance (e.g. Euclidean) measured between observations}
\citep{foote1993contributions,Foote29111996}
\hl{, a more direct approximation of the hyper volume }
\citep{cornwell2006trait,DonohueDim}
\hl{, or many more }
\citep[e.g.][]{Navarrro2003MDA}.

Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

@@

This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
\hl{The first argument is the matrix}
defining the multidimensional space and a list of elements to group into different subsamples.
The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example, 
\hl{see below}
) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
